---
title: "Early Childhood Program Evaluation: A Regression Analysis of WIC and AFDC Effects on Child Academic Achievement"
author: "Liqi Liang"
date: "2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This project uses data from the Child Development Supplement of the Panel Study of Income Dynamics (PSID-CDS) to evaluate how participation in federally funded early childhood programs — specifically WIC (Women, Infants, and Children Nutrition Program) and AFDC (Aid to Families with Dependent Children) — relates to children's academic achievement. The analysis proceeds in three stages:

- **Part 1:** Multiple regression examining the individual and joint effects of WIC and AFDC participation on children's reading achievement
- **Part 2:** Regression assumption diagnostics (linearity, homoscedasticity, normality) and outlier corrections, applied to a model predicting math achievement
- **Part 3:** Interaction effects — testing whether the effect of WIC participation on math achievement is moderated by family income, race, and child age

All analyses control for child age, family income, birth weight status, and home environment quality (HOME97).

---

```{r load-libraries, message=FALSE, warning=FALSE}
rm(list = ls(all.names = TRUE))
library(stargazer)
library(knitr)
```

---

## Part 1: WIC and AFDC Effects on Reading Achievement

### Data Preparation

```{r part1-data}
data1 <- read.csv("good.csv")
subset_good <- c("readss97", "WICpreg", "AFDCpreg", "AGE97", "faminc97", "bthwht", "HOME97")
good <- data1[, subset_good]
good <- na.omit(good)
nrow(good)
```

### Descriptive Statistics

```{r part1-descriptives}
summary(good)
sapply(good, function(x) round(sd(x), 2))
```

### Correlation Matrix

```{r part1-correlation}
round(cor(good), 4)
```

### Regression Models

Three nested OLS models are estimated:

- **Model 1a:** WIC participation only (plus controls)
- **Model 1b:** AFDC participation only (plus controls)
- **Model 1c:** Both WIC and AFDC together (plus controls)

```{r part1-models}
model1_WIC <- lm(readss97 ~ WICpreg + AGE97 + faminc97 + bthwht + HOME97, data = good)
summary(model1_WIC)

model2_AFDC <- lm(readss97 ~ AFDCpreg + AGE97 + faminc97 + bthwht + HOME97, data = good)
summary(model2_AFDC)

model3_both <- lm(readss97 ~ WICpreg + AFDCpreg + AGE97 + faminc97 + bthwht + HOME97, data = good)
summary(model3_both)
```

```{r part1-table, results='asis'}
stargazer::stargazer(model1_WIC, model2_AFDC, model3_both, type = "html",
                     title = "OLS Regression: Effects of WIC and AFDC on Reading Achievement")
```

### Partial F-Tests: Unique Contributions

```{r part1-anova}
# Does adding WIC improve fit beyond AFDC alone?
anova(model1_WIC, model3_both)
# Does adding AFDC improve fit beyond WIC alone?
anova(model2_AFDC, model3_both)
```

---

## Part 2: Assumption Diagnostics and Outlier Corrections

This section evaluates whether the OLS assumptions hold for a model predicting **math achievement** (mathraw97), and applies corrections where violations are found.

### Data Preparation

Family income is log-transformed to address right skew (negative incomes set to NA).

```{r part2-data}
data1 <- read.csv("good.csv")
data1$loginc97 <- ifelse(data1$faminc97 > 0, log(data1$faminc97), NA)
subset_good <- c("mathraw97", "AGE97", "loginc97", "bthwht", "WICpreg", "HOME97")
good <- data1[, subset_good]
good <- na.omit(good)
nrow(good)
```

```{r part2-descriptives}
summary(good)
sapply(good, function(x) round(sd(x), 2))
round(cor(good), 4)
```

### Base Model

```{r part2-model1}
model1 <- lm(mathraw97 ~ AGE97 + loginc97 + bthwht + WICpreg, data = good)
summary(model1)
```

### Assumption Diagnostics

**Linearity** — Scatterplots of mathraw97 against each predictor:

```{r part2-linearity}
par(mfrow = c(2, 2))
plot(good$AGE97, good$mathraw97, main = "Mathraw97 vs AGE97")
abline(lm(mathraw97 ~ AGE97, data = good), col = "skyblue")
plot(good$loginc97, good$mathraw97, main = "Mathraw97 vs loginc97")
abline(lm(mathraw97 ~ loginc97, data = good), col = "skyblue")
plot(good$bthwht, good$mathraw97, main = "Mathraw97 vs bthwht")
abline(lm(mathraw97 ~ bthwht, data = good), col = "skyblue")
plot(good$WICpreg, good$mathraw97, main = "Mathraw97 vs WICpreg")
abline(lm(mathraw97 ~ WICpreg, data = good), col = "skyblue")
```

**Homoscedasticity** — Residuals vs. fitted values:

```{r part2-homoscedasticity}
plot(model1$fitted.values, model1$residuals, main = "Residuals vs Fitted")
abline(h = 0, col = "deeppink")
lines(lowess(model1$fitted.values, model1$residuals), col = "blue")
```

**Normality of Residuals:**

```{r part2-normality}
par(mfrow = c(1, 2))
hist(model1$residuals, breaks = 20, main = "Histogram of Residuals")
qqnorm(model1$residuals)
qqline(model1$residuals, col = "blue")
```

### Outlier Diagnostics

**Studentized Residuals** (threshold: ±3):

```{r part2-outliers-residuals}
good$index <- 1:nrow(good)
good$rstudent <- rstudent(model1)

plot(good$rstudent, xlab = "Index", ylab = "Studentized residuals", pch = 20)
abline(h = 0, col = "deeppink")
abline(h = c(2, -2), col = "blue")
abline(h = c(3, -3), col = "lightgreen")

outliers_residuals <- subset(good, abs(good$rstudent) > 3)
print(outliers_residuals$index)
```

**Leverage Values** (threshold: 0.03):

```{r part2-outliers-leverage}
cutoff_leverage <- (2 * (ncol(good) - 1)) / nrow(good)
good$leverage <- hatvalues(model1)

plot(good$leverage, xlab = "Index", ylab = "Leverage", pch = 19)
abline(h = cutoff_leverage, col = "green")
abline(h = 0.03, col = "deeppink")

outliers_leverage <- subset(good, good$leverage > 0.03)
print(outliers_leverage$index)
```

**Cook's Distance** (threshold: 4/n):

```{r part2-outliers-cooks}
cutoff_cooksd <- 4 / nrow(good)
good$cooksd <- cooks.distance(model1)

influential_points <- subset(good, good$cooksd > cutoff_cooksd)
print(influential_points$index)

hist(good$cooksd[good$cooksd > cutoff_cooksd],
     main = "Cook's Distance for Influential Points",
     xlab = "Cook's Distance",
     breaks = 10)

cook_quantiles <- round(quantile(good$cooksd[good$cooksd > cutoff_cooksd], probs = seq(0, 1, 0.05)), 4)
print(cook_quantiles)
```

### Corrected Model and HOME97 Inclusion

Observations flagged by any of the three outlier criteria are removed. A third model adds HOME97 (home environment quality) to assess its incremental contribution.

```{r part2-corrections}
good_clean <- good[!(good$index %in% c(outliers_residuals$index,
                                        outliers_leverage$index,
                                        influential_points$index)), ]
nrow(good_clean)

model2_corrected <- lm(mathraw97 ~ AGE97 + loginc97 + bthwht + WICpreg, data = good_clean)
summary(model2_corrected)

model3 <- lm(mathraw97 ~ AGE97 + loginc97 + bthwht + WICpreg + HOME97, data = good_clean)
```

```{r part2-table, results='asis'}
stargazer::stargazer(model1, model2_corrected, model3, type = "html",
                     title = "OLS Regression: Math Achievement — Base, Corrected, and Extended Models",
                     column.labels = c("Base", "Outliers Removed", "+ HOME97"))
```

---

## Part 3: Interaction Effects

This section tests whether the effect of WIC participation on math achievement is moderated by family income, race, and child age.

### Data Preparation

Race is recoded as a binary variable (0 = White, 1 = non-White; observations with CHRACE = 9 excluded as missing).

```{r part3-data}
data1 <- read.csv("good.csv")
data1$loginc97 <- ifelse(data1$faminc97 > 0, log(data1$faminc97), NA)
subset_good <- c("WICpreg", "CHRACE", "mathraw97", "AGE97", "loginc97", "bthwht", "HOME97")
good <- data1[, subset_good]
good$RACE <- ifelse(good$CHRACE == 9, NA,
             ifelse(good$CHRACE == 1, 0,
             ifelse(good$CHRACE == 2, 1, NA)))
good <- na.omit(good)
nrow(good)
```

```{r part3-descriptives}
summary(good)
sapply(good, function(x) round(sd(x), 2))
round(cor(good), 4)
```

### Models

```{r part3-models}
# Base model (no interactions)
model_base <- lm(mathraw97 ~ loginc97 + RACE + AGE97 + WICpreg + bthwht + HOME97, data = good)

# Interaction models
model_1 <- lm(mathraw97 ~ loginc97 + RACE + AGE97 + WICpreg + bthwht + HOME97 + WICpreg:loginc97, data = good)
model_2 <- lm(mathraw97 ~ loginc97 + RACE + AGE97 + WICpreg + bthwht + HOME97 + WICpreg:RACE, data = good)
model_3 <- lm(mathraw97 ~ loginc97 + RACE + AGE97 + WICpreg + bthwht + HOME97 + WICpreg:AGE97, data = good)

# Full model with all three interaction terms
model_full <- lm(mathraw97 ~ loginc97 + RACE + AGE97 + WICpreg + bthwht + HOME97 +
                   WICpreg:loginc97 + WICpreg:RACE + WICpreg:AGE97, data = good)
```

### ANOVA Model Comparisons

```{r part3-anova}
aov_loginc97 <- anova(model_base, model_1)
aov_RACE     <- anova(model_base, model_2)
aov_AGE97    <- anova(model_base, model_3)

print("ANOVA: WICpreg × loginc97")
print(aov_loginc97)
print("ANOVA: WICpreg × RACE")
print(aov_RACE)
print("ANOVA: WICpreg × AGE97")
print(aov_AGE97)
```

```{r part3-table, results='asis'}
stargazer::stargazer(model_base, model_1, model_2, model_3, model_full,
                     type = "html",
                     title = "Regression Models with Interaction Effects",
                     column.labels = c("Base", "× Income", "× Race", "× Age", "Full"))
```
